Latent Predicate Networks: unsupervised learning of probabilistic programs with context sensitive grammars

### 2014-11-06 drafts/nips_workshop/nips_workshop.tex
significant cuts for length of NIPS paper

\title{Latent Predicate Networks: Concept Learning with Probabilistic Context-Sensitive Grammars}

\author{Eyal Dechter\thanks{{\tt \{edechter, rule, jbt\} @mit.edu}}}
\author{Joshua Rule}
\author{Joshua B. Tenenbaum}
\affil{Department of Brain and Cognitive Sciences, MIT}

@@

\begin{abstract}
  For humans, learning abstract concepts and learning language go hand
  in hand: we acquire abstract knowledge primarily through linguistic
  experience, and acquiring abstract concepts is a crucial step in
  learning the meanings of linguistic expressions. Number knowledge is
  a case in point: we largely acquire concepts such as seventy-three
  through linguistic means, and we can only know what the sentence
  ``seventy-three is more than twice as big as thirty-one" means if we
  can grasp the meanings of its component number words. How do we
  begin to solve this problem? One approach is to estimate the
  distribution from which sentences are drawn, and, in doing so, infer
  the latent concepts and relationships that best explain those
  sentences. We present early work on a learning framework called
  Latent Predicate Networks (LPNs) which learns concepts by inferring
  the parameters of probabilistic context-sensitive grammars over
  sentences.  We show that for a small fragment of sentences
  expressing relationships between English number words, we can use
  hierarchical Bayesian inference to learn LPNs that can answer simple
  queries about previously unseen relationships within this
  domain. These generalizations demonstrate LPNs' promise as a tool
  for learning and representing conceptual knowledge in language.
\end{abstract}

@@

\subsection{Generic Architecture}

An LPN is a PRCG with the following generic architecture: the
\emph{axiom} of an LPN is a unary predicate $S$, and it has $K$
densely connected binary predicates $\{A_k\}_{k=1}^K$ (Figure
\ref{fig:architecture}). For each latent predicate $A_k$, we define
rules such that $A_k(w_a,w_b)$ is true for each possible pair of
terminals, $w_a,w_b \in T$. $T$ may include the empty string $\epsilon$, but we
do not allow latent predicates of the form $A_k(\epsilon, \epsilon)$:
see Section~\ref{sec:implementation}. We also define rules such that
$A_k(XY,UV)$ is true for each possible pair of predicates, $A_i,A_j$,
and every possible ordering of the variables $X,Y,U$ and $V$ across
$A_i$ and $A_j$. Finally, we define $S(XY)$ to be the concatenation of
the two arguments of $A_1(X,Y)$.

@@

\begin{figure}[t]
	\centering
	\begin{minipage}[b]{0.8\linewidth}
1) A(a, a) :: 0.25. \\
2) A(b, b) :: 0.3. \\
3) A(X a, Y a) $\leftarrow$ A(X, Y) :: 0.25. \\
4) A(X b, Y b) $\leftarrow$ A(X, Y) :: 0.2.
		\subcaption{}
		\label{fig:grammar}
	\end{minipage}
	%\rule[0pt]{0.8\linewidth}{.5pt}
	\begin{minipage}[b]{0.8\linewidth}
        \fontsize{9}{10.5}\selectfont\ttfamily
		\begin{verbatim}
values(`A', [1, 2, 3, 4],
    [0.25, 0.30, 0.25, 0.20]).

reduce(`A'-[[a],[a]],1).
reduce(`A'-[[b],[b]],2).
reduce(`A'-[A2,B2],3) :- lpn(`A'-[X,Y]),
    append(X,[a],A2), append(Y,[a],B2).
reduce(`A'-[A2,B2],4) :- lpn(`A'-[X,Y]),
    append(X,[b],A2), append(Y,[b],B2).

lpn(P-IN) :- reduce(P-IN,V), msw(P,V).
		\end{verbatim}
		\subcaption{}
		\label{fig:prism}
	\end{minipage}
	\caption{Possible encodings of the 2-copy language $\{ww\,|\, w \in \{a,b\}^+\}$ as (\subref{fig:grammar}) an LPN, (\subref{fig:prism}) a PRISM program.}
	\label{fig:copy}
\end{figure}

LPNs can be encoded as a restricted subclass of PRISM programs; this
is very similar to how PCFGs are encoded in
PRISM~\cite{DBLP:conf/cl/2000}. See Figure~\ref{fig:copy} for an
example. A general probabilistic logic programming system based on
PROLOG, PRISM provides built-in predicates for probabilistic execution
and Bayesian inference over logic programs with stochastic
choices. There are several restrictions placed on PRISM programs to
maintain the validity of their probabilistic interpretation. Most
importantly, derivations of a probabilistic predicate cannot contain
cycles. Because we disallow $A_i(\epsilon,\epsilon)$ as a valid
clause, every term in the body of a clause has shorter arguments than
the head, giving acyclic and finite derivations.

@@

\begin{table}[h]
  \begin{subtable}[b]{0.5\linewidth}
    \begin{tabular}{>{\tiny} l >{\tiny} l}
      \multicolumn{2}{>{\tiny}c}{$S$(X Y) $\leftarrow$ $A_1$(X, Y) : 1.0000} \\
      \multicolumn{2}{>{\tiny}c}{$A_1$(X Y, U V) $\leftarrow$ $A_2$(X, U), $A_3$(V, Y) : 0.5002} \\
      \multicolumn{2}{>{\tiny}c}{$A_1$(X Y, U V) $\leftarrow$ $A_3$(Y, V), $A_4$(X, U) : 0.3428} \\
      \multicolumn{2}{>{\tiny}c}{$A_1$(X Y, U V) $\leftarrow$ $A_1$(V, Y), $A_4$(X, U) : 0.0796} \\
      \multicolumn{2}{>{\tiny}c}{$A_1$(X Y, U V) $\leftarrow$ $A_1$(Y, V), $A_2$(X, U) : 0.0712} \\
      \multicolumn{2}{>{\tiny}c}{$A_1$(X Y, U V) $\leftarrow$ $A_2$(Y, X), $A_3$(V, U) : 0.0021} \\
      \multicolumn{2}{>{\tiny}c}{$A_1$(X Y, U V) $\leftarrow$ $A_1$(V, Y), $A_2$(X, U) : 0.0013} \\
      \multicolumn{2}{>{\tiny}c}{$A_1$(X Y, U V) $\leftarrow$ $A_2$(V, X), $A_3$(Y, U) : 0.0008} \\
      \multicolumn{2}{>{\tiny}c}{$A_1$(X Y, U V) $\leftarrow$ $A_1$(X, U), $A_2$(V, Y) : 0.0008} \\
      \multicolumn{2}{>{\tiny}c}{$A_1$(X Y, U V) $\leftarrow$ $A_1$(X, U), $A_2$(Y, V) : 0.0008} \\
      \multicolumn{2}{>{\tiny}c}{$A_1$(X Y, U V) $\leftarrow$ $A_1$(X, U), $A_4$(Y, V) : 0.0004} \\
      $A_2$(before, comes) : 0.7316 & $A_4$(after, comes) : 0.9990  \\
      $A_3$(one, two) : 0.3993 &
      $A_3$(two, three) : 0.2063 \\
      $A_3$(three, four) : 0.1093 &
      $A_3$(four, five) : 0.0734 \\
      $A_3$(five, six) : 0.0502 &
      $A_3$(six, seven) : 0.0355 \\
      $A_3$(eight, nine) : 0.0290 &
      $A_3$(seven, eight) : 0.0271 \\
      $A_2$(fifty, fifty) : 0.0375 &
      $A_2$(thirty, thirty) : 0.0361 \\
      $A_2$(eighty, eighty) : 0.0339 &
      $A_3$(null, one) : 0.0231 \\
      $A_2$(forty, forty) : 0.0332 &
      $A_2$(twenty, twenty) : 0.0310 \\
      $A_2$(seventy, seventy) : 0.0296 &
      $A_2$(sixty, sixty) : 0.0274 \\
      $A_2$(ninety, ninety) : 0.0260 &
      $A_3$(eighteen, nineteen) : 0.0064 \\
      $A_3$(sixteen, seventeen) : 0.0049 &
      $A_3$(eleven, twelve) : 0.0044 \\
      $A_3$(nine, ten) : 0.0044 &
      $A_3$(thirteen, fourteen) : 0.0044 \\
      $A_3$(fourteen, fifteen) : 0.0039 &
      $A_3$(ten, eleven) : 0.0034 \\
      $A_2$(null, fifty) : 0.0043 &
      $A_3$(eighty, null) : 0.0030 \\
      $A_3$(seventeen, eighteen) : 0.0030 &
      $A_3$(nine, sixty) : 0.0025 \\
      $A_2$(nine, seventy) : 0.0029 &
      $A_3$(nine, forty) : 0.0020 \\
      $A_2$(null, thirty) : 0.0022 &
      $A_2$(nine, ninety) : 0.0022 \\
      $A_3$(twelve, thirteen) : 0.0015 &
      $A_3$(fifteen, sixteen) : 0.0015 \\
      $A_3$(null, comes) : 0.0010 &
      $A_2$(twenty, after) : 0.0014 \\
      $A_2$(sixty, before) : 0.0007 &
      $A_3$(nineteen, comes) : 0.0005 \\
      $A_4$(nine, thirty) : 0.0010 & \\
    \end{tabular}
    \caption{The 4 predicate LPN trained to model sentences in the number domain. Rules with insignificant weights are removed. This LPN generates the completions in Table \ref{tab:results}.}
    \label{tab:grammar}
  \end{subtable}
  \begin{subtable}[b]{0.5\linewidth}
    \begin{tabular}{>{\footnotesize} l >{\footnotesize} l}
      Question & $K=4$ \\ \hline
      after twenty comes \underline{\hspace{1cm}}? & twenty one \checkmark \\
      after forty five comes \underline{\hspace{1cm}}? & forty six \checkmark \\
      after forty seven comes \underline{\hspace{1cm}}? & forty eight  \checkmark \\
      after forty nine comes \underline{\hspace{1cm}}? & forty ten $\times$ \\
      after fifty nine comes \underline{\hspace{1cm}}? & fifty ten $\times$ \\
      after sixty one comes \underline{\hspace{1cm}}? & sixty two \checkmark \\
      after sixty three comes \underline{\hspace{1cm}}? & sixty four \checkmark \\
      after sixty four comes \underline{\hspace{1cm}}? & sixty five \checkmark \\
      after sixty five comes \underline{\hspace{1cm}}? & sixty six \checkmark \\
      after sixty nine comes \underline{\hspace{1cm}}? & sixty ten $\times$ \\
      after seventy three comes \underline{\hspace{1cm}}? & seventy four \checkmark \\
      after seventy nine comes \underline{\hspace{1cm}}? & seventy ten $\times$ \\
      after ninety five comes \underline{\hspace{1cm}}? & ninety six \checkmark \\
      before twenty three comes \underline{\hspace{1cm}}? & twenty two \checkmark \\
      before thirty comes \underline{\hspace{1cm}}? & thirty eighty $\times$ \\
      before thirty eight comes \underline{\hspace{1cm}}? & thirty seven \checkmark \\
      before forty one comes \underline{\hspace{1cm}}? & forty \checkmark \\
      before fifty three comes \underline{\hspace{1cm}}? & fifty two \checkmark \\
      before sixty eight comes \underline{\hspace{1cm}}? & sixty seven \checkmark \\
      before seventy two comes \underline{\hspace{1cm}}? & seventy one \checkmark \\
      before seventy three comes \underline{\hspace{1cm}}? & seventy two \checkmark \\
      before eighty five comes \underline{\hspace{1cm}}? & eighty four \checkmark \\
      before ninety two comes \underline{\hspace{1cm}}? & ninety one \checkmark \\
      before ninety three comes \underline{\hspace{1cm}}? & ninety two \checkmark \\
      before ninety five comes \underline{\hspace{1cm}}? &ninety four \checkmark \\
    \end{tabular}
    \caption{Viterbi completions of held-out sentences to evaluate an LPN with 4 latent predicates in the number domain.}
    \label{tab:results}
  \end{subtable}
\end{table}

@@

\section{Discussion}

Though simple, the knowledge learned by LPNs in the number
domain we explore in this paper goes beyond the expressive capacity of
PCFGs and HMMs, and yet it avoids intractability by using a restricted
formalism. The human ability to learn abstract concepts must also rely
on such a compromise, and it remains to be seen to what extent LPNs
can model this ability.

This work leaves many open questions. What domains of concepts are
amenable to being represented and learned by LPNs? Can multiple and
varied systems of concepts be learned at once? Perhaps most
importantly, are there generic architectures for LPNs that allow a
larger number of predicates without an insurmountable blow-up in the
number of rules? One idea which we are pursuing is using layered
connectivity, much like neural networks, to limit the number of rules
in which each predicate participates.

Many of the algorithms and formalisms used here were originally
developed for use in other closely related areas, including logic
programming and semantic parsing. Semantic parsing, in particular,
seems especially related to the challenge we face here of jointly
learning meaning and structure from
sentences~\cite{berant2013semantic, liang2013learning, kwiatkowski2010inducing, poon2009unsupervised}. Semantic parsing,
however, tends to frame this challenge slightly differently than we do
here, namely, by asking how utterances can be mapped to an explicit
internal logical language. By contrast, we focus here on systems where
meaning and structure seem inseparable. Understanding how these two
approaches relate and inform one another is an interesting and open
question.

\subsubsection*{Acknowledgments}

We should probably include acknowledgments for this paper.

\subsubsection*{References}

>>>>

\title{Latent Predicate Networks: unsupervised learning of  probabilistic programs with context sensitive grammars}

\author{Eyal Dechter\thanks{{\tt \{edechter, rule, jbt\} @mit.edu}}}
\author{Joshua Rule}
\author{Joshua B. Tenenbaum}
\affil{{\normalsize Department of Brain and Cognitive Sciences, MIT}}
\date{}

@@

\begin{abstract}

%%   - learning both abstract concepts and language presents an interesting chicken and egg problem
%% 
%%   - we can model this sort of learning a conceptual grammar for sentences
%% 
%%   - importantly, this grammar must be probabilistic to accomodate noisy data and represent our uncertainty about a concept's meaning
%% 
%%   - Here, we present early work on LPNs, showing them to be the solution we want
%% 
%%   - Crucially, we show how, for certain domains, learning language and abstract concepts can be done simultaneously
%% 
%%   - We also show that this approach naturally represents sentences as probabilistic conceptual programs.
%% 
%% 
%%   While a large set of abstract concepts are necessary to understand
%%   the utterances of a language, very few of those concepts are
%%   initially available to children. Instead, they must be learned,
%%   primarily {\it via} language. One approach to understanding this
%%   joint learning problem is to model sentences as the output of
%%   probabilistic programs constructed
%% 
%% 
%% 
%%   A large set of abstract concepts are necessary to understand the
%%   utterances of a language, and at the same time, many of those
%%   abstract concepts are acquired primarily {\it via} linguistic
%%   experience. One way we might approach this sort of simultaneous
%%   learning problem is to infer the distribution over sentences, where
%%   the latent structure of this distributi
%% 
  For humans, learning abstract concepts and learning language go hand
  in hand: we acquire much of our abstract knowledge through
  linguistic experience, but the same abstract concepts are often
  crucial to interpreting linguistic expressions. How can we do both
  simultaneously? One approach is to estimate the distribution from
  which sentences are drawn, and, in doing so, infer the latent
  concepts and relationships that best explain those sentences. We
  present early work on a learning framework called Latent Predicate
  Networks (LPNs) which learns concepts by inferring the parameters of
  probabilistic context-sensitive grammars over sentences.  We show
  that for a small fragment of sentences expressing relationships
  between English number words, we can use online hierarchical Bayesian
  inference to learn LPNs that can answer simple queries about
  previously unseen relationships within this domain. These
  generalizations demonstrate LPNs' promise as a tool for learning and
  representing conceptual knowledge in language.
\end{abstract}

@@

@@

LPNs can be encoded as a restricted subclass of PRISM programs; this
is very similar to how PCFGs are encoded in
PRISM~\cite{DBLP:conf/cl/2000}. A general probabilistic logic
programming system based on PROLOG, PRISM provides built-in predicates
for probabilistic execution and Bayesian inference over logic programs
with stochastic choices. There are several restrictions placed on
PRISM programs to maintain the validity of their probabilistic
interpretation. Most importantly, derivations of a probabilistic
predicate cannot contain cycles. Because we disallow
$A_i(\epsilon,\epsilon)$ as a valid clause, every term in the body of
a clause has shorter arguments than the head, giving acyclic and
finite derivations.

@@

  \begin{table}
    \centering
    \begin{tabular}{>{\tiny} l >{\tiny} l >{\tiny} l >{\tiny} l }
      \multicolumn{2}{>{\tiny}l}{$S$(X Y) $\leftarrow$ $A_1$(X, Y) : 1.0000} &      $A_2$(before, comes) : 0.7316 & $A_4$(after, comes) : 0.9990  \\
      \multicolumn{2}{>{\tiny}l}{$A_1$(X Y, U V) $\leftarrow$ $A_2$(X, U), $A_3$(V, Y) : 0.5002} & $A_3$(one, two) : 0.3993 & $A_3$(two, three) : 0.2063 \\
      \multicolumn{2}{>{\tiny}l}{$A_1$(X Y, U V) $\leftarrow$ $A_3$(Y, V), $A_4$(X, U) : 0.3428} & $A_3$(three, four) : 0.1093 & $A_3$(four, five) : 0.0734 \\
      \multicolumn{2}{>{\tiny}l}{$A_1$(X Y, U V) $\leftarrow$ $A_1$(V, Y), $A_4$(X, U) : 0.0796} & $A_3$(five, six) : 0.0502 & $A_3$(six, seven) : 0.0355 \\
      \multicolumn{2}{>{\tiny}l}{$A_1$(X Y, U V) $\leftarrow$ $A_1$(Y, V), $A_2$(X, U) : 0.0712} & $A_3$(eight, nine) : 0.0290 & $A_3$(seven, eight) : 0.0271 \\
      \multicolumn{2}{>{\tiny}l}{$A_1$(X Y, U V) $\leftarrow$ $A_2$(Y, X), $A_3$(V, U) : 0.0021} & $A_2$(fifty, fifty) : 0.0375 & $A_2$(thirty, thirty) : 0.0361 \\
      \multicolumn{2}{>{\tiny}l}{$A_1$(X Y, U V) $\leftarrow$ $A_1$(V, Y), $A_2$(X, U) : 0.0013} & $A_2$(eighty, eighty) : 0.0339 & $A_3$(null, one) : 0.0231 \\
      \multicolumn{2}{>{\tiny}l}{$A_1$(X Y, U V) $\leftarrow$ $A_2$(V, X), $A_3$(Y, U) : 0.0008} & $A_2$(forty, forty) : 0.0332 & $A_2$(twenty, twenty) : 0.0310 \\
      \multicolumn{2}{>{\tiny}l}{$A_1$(X Y, U V) $\leftarrow$ $A_1$(X, U), $A_2$(V, Y) : 0.0008} & $A_2$(seventy, seventy) : 0.0296 & $A_2$(sixty, sixty) : 0.0274 \\
      \multicolumn{2}{>{\tiny}l}{$A_1$(X Y, U V) $\leftarrow$ $A_1$(X, U), $A_2$(Y, V) : 0.0008} & $A_2$(ninety, ninety) : 0.0260 & $A_3$(eighteen, nineteen) : 0.0064 \\
      \multicolumn{2}{>{\tiny}l}{$A_1$(X Y, U V) $\leftarrow$ $A_1$(X, U), $A_4$(Y, V) : 0.0004} & $A_3$(sixteen, seventeen) : 0.0049 & $A_3$(eleven, twelve) : 0.0044 \\
      $A_3$(nine, ten) : 0.0044 & $A_3$(thirteen, fourteen) : 0.0044 & $A_3$(fourteen, fifteen) : 0.0039 & $A_3$(ten, eleven) : 0.0034 \\
      $A_2$(null, fifty) : 0.0043 & $A_3$(eighty, null) : 0.0030 & $A_3$(seventeen, eighteen) : 0.0030 & $A_3$(nine, sixty) : 0.0025 \\
      $A_2$(nine, seventy) : 0.0029 & $A_3$(nine, forty) : 0.0020 & $A_2$(null, thirty) : 0.0022 & $A_2$(nine, ninety) : 0.0022 \\
      $A_3$(twelve, thirteen) : 0.0015 & $A_3$(fifteen, sixteen) : 0.0015 & $A_3$(null, comes) : 0.0010 & $A_2$(twenty, after) : 0.0014 \\
      $A_2$(sixty, before) : 0.0007 & $A_3$(nineteen, comes) : 0.0005 & $A_4$(nine, thirty) : 0.0010 & \\
    \end{tabular}
    \caption{The 4 predicate LPN trained to model sentences in the number domain. Rules with insignificant weights are removed. This LPN generates the completions in Table \ref{tab:results}.}
    \label{tab:grammar}
  \end{table}

\begin{figure}
    \begin{subfigure}{0.45\linewidth}
      \begin{tabular}{>{\footnotesize} l >{\footnotesize} l}
        Question & $K=4$ \\ \hline
        e.g. after twenty comes \underline{\hspace{1cm}}? & twenty one \checkmark \\
        e.g. after forty five comes \underline{\hspace{1cm}}? & forty six \checkmark \\
        e.g. after forty seven comes \underline{\hspace{1cm}}? & forty eight  \checkmark \\
        e.g. after forty nine comes \underline{\hspace{1cm}}? & forty ten $\times$ \\
        e.g. after fifty nine comes \underline{\hspace{1cm}}? & fifty ten $\times$ \\
        e.g. after sixty one comes \underline{\hspace{1cm}}? & sixty two \checkmark \\
        %% after sixty three comes \underline{\hspace{1cm}}? & sixty four \checkmark \\
        %% after sixty four comes \underline{\hspace{1cm}}? & sixty five \checkmark \\
        %% after sixty five comes \underline{\hspace{1cm}}? & sixty six \checkmark \\
        %% after sixty nine comes \underline{\hspace{1cm}}? & sixty ten $\times$ \\
        %% after seventy three comes \underline{\hspace{1cm}}? & seventy four \checkmark \\
        %% after seventy nine comes \underline{\hspace{1cm}}? & seventy ten $\times$ \\
        %% after ninety five comes \underline{\hspace{1cm}}? & ninety six \checkmark \\
        %% before twenty three comes \underline{\hspace{1cm}}? & twenty two \checkmark \\
        %% before thirty comes \underline{\hspace{1cm}}? & thirty eighty $\times$ \\
        %% before thirty eight comes \underline{\hspace{1cm}}? & thirty seven \checkmark \\
        %% before forty one comes \underline{\hspace{1cm}}? & forty \checkmark \\
        %% before fifty three comes \underline{\hspace{1cm}}? & fifty two \checkmark \\
        %% before sixty eight comes \underline{\hspace{1cm}}? & sixty seven \checkmark \\
        %% before seventy two comes \underline{\hspace{1cm}}? & seventy one \checkmark \\
        %% before seventy three comes \underline{\hspace{1cm}}? & seventy two \checkmark \\
        %% before eighty five comes \underline{\hspace{1cm}}? & eighty four \checkmark \\
        %% before ninety two comes \underline{\hspace{1cm}}? & ninety one \checkmark \\
        %% before ninety three comes \underline{\hspace{1cm}}? & ninety two \checkmark \\
        %% before ninety five comes \underline{\hspace{1cm}}? &ninety four \checkmark \\
      \end{tabular}
      \caption{Viterbi completions of held-out sentences to evaluate an LPN with 4 latent predicates in the number domain.}
      \label{tab:results}
  \end{subfigure}
  \hfill
  \begin{subfigure}{0.45\linewidth}
    \rule{6cm}{2.5cm}
    \caption{PUT HELD OUT LL FIGURE HERE}
    \label{fig:heldoutLL}
  \end{subfigure}
\end{figure}

@@

\subsubsection*{Acknowledgments?}
